---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ineq

<!-- badges: start -->
<!-- badges: end -->

The goal of `ineq` is to create a set of functions to compute various measures of economic inequality. You provide the data, choose the relevant function and the package will compute the measures.  

## Basic Measures of Economic Inequality

The three main measures of inequality that is created by this package are the Lorenz curve, the coefficient of variation (CV) and the Gini coefficient (Gini). 

Consider a society with $j = 1, 2, \ldots, m$ income groups,  

- where $y_j$ is the mean income level in group $j$, 
- the groups are arranged in increasing order of mean income, i.e., $y_1 \leq y_2, \ldots, \leq y_m$, and
- the number of persons in the groups are given by $n_1 \leq n_2, \ldots, \leq n_m$.

Note that the total number of persons in this society is given by $$n = n_1 + n_2 + \cdots + n_m$$ and the total income level in this society is given by $$y = n_1y_1 + n_2 y_2 + \cdots + n_my_m$$

### Lorenz Curve

The most important measure of inequality is the Lorenz curve, which is a plot of the cumulative income share against the cumulative population share. To construct this plot, we first create a table of the two variables, the cumulative income share and the cumulative population share.

Let $w_j = n_j/n$ denote the population share and $s_j=n_jy_j/y$ denote the income share of the $j$-th group. Then the cumulative population share is given by $cw_j = \sum_{i=1}^{j}w_i$ and the cumulative income share is given by $cs_j = \sum_{i=1}^{j}s_i$. Now consider the plane where $cw$ is measured on the horizontal axis and $cs$ is measured on the vertical axis. A line plot of $cs$ against $cw$ is called the *Lorenz curve*. 

Since both $cs$ and $cw$ lies between $0$ and $1$, the Lorenz curve fits within a square whose side is of length $1$ (or 100, if the shares are measured in percentages). If the income distribution is perfectly equal, then each group's income share is the same. In that case, the Lorenz curve is the $45$-degree line through the origin. As distributions of income become more unequal, the Lorenz curve bows out and away from the $45$-degree line.

### Coefficient of Variation

The mean income level is given by $$\mu = \frac{1}{n} \sum_{j=1}^{m} n_j y_j = \sum_{j=1}^{m} w_j y_j$$
and the variance of the income level is given by $$\sigma^2 = \frac{1}{n} \sum_{j=1}^{m} n_j \left(y_j-\mu \right)^2 = \sum_{j=1}^{m} w_j \left(y_j-\mu \right)^2.$$

The second measure of inequality is the coefficient of variation: $$CV=\sigma/\mu$$.

### Gini coefficient

The third measure of inequality is the Gini coefficient: $$G = \frac{1}{2n^2 \mu} \sum_{j=1}^{m} \sum_{k=1}^{m} n_j n_k |y_j-y_k| = \frac{1}{2 \mu} \sum_{j=1}^{m} \sum_{k=1}^{m} w_j w_k |y_j-y_k|.$$


### Theil's first measure 

The third measure of inequality is Theil's first measure: $$T = \sum_{j=1}^{m} \frac{n_j}{n}  \frac{y_j}{\mu} \ln \left( \frac{y_j}{\mu} \right).$$


### Using data on population and income shares

In many situations, researchers have access *solely* to data on population shares and income shares of each group, i.e. $w_j$ and $s_j$, for $j=1,2, \ldots, m$, but they do not have access to data on mean income levels in each group. For example, a researcher might only know the income shares of each quintile of the income distribution. How do we compute the measures of economic inequality in this case?

The Lorenz curve requires data on cumulative population and income shares. Using $w_j$ and $s_j$, we can construct these cumulative shares and construct the Lorenz curve. 

The CV and Gini coefficient can also be expressed in terms of population and income shares. To see this, note that since $s_j=n_jy_j/y=n_jy_j/n\mu=w_jy_j/\mu$, we get $y_j=(s_j/w_j)\mu$. 

Using this, we see that the variance of the income distribution is given by

$$\sigma^2 = \sum_{j=1}^{m} w_j \left(y_j-\mu \right)^2 = \sum_{j=1}^{m} \mu^2 w_j \left(\frac{s_j}{w_j}-1 \right)^2 = \mu^2 \sum_{j=1}^{m}  w_j \left(\frac{s_j}{w_j}-1 \right)^2.$$ 

Hence, the coefficient is variation is given by
$$CV=\frac{\sigma}{\mu}=\sum_{j=1}^{m}  w_j \left(\frac{s_j}{w_j}-1 \right)^2$$

In a similar way, using $y_j=(s_j/w_j)\mu$, we see that the Gini coefficient is given by $$G = \frac{1}{2n^2 \mu} \sum_{j=1}^{m} \sum_{k=1}^{m} n_j n_k |y_j-y_k| = \frac{1}{2 \mu} \sum_{j=1}^{m} \sum_{k=1}^{m} w_j w_k |(s_j/w_j)-(s_k/w_k)|$$

In the same way, Theil's first measure can re-written as: $$T = \sum_{j=1}^{m} \frac{n_j}{n}  \frac{y_j}{\mu} \ln \left( \frac{y_j}{\mu} \right) = \sum_{j=1}^{m} s_j \ln \left( \frac{y_j}{\mu} \right).$$


## The functions in the package

The following functions are available right now:  

- `ineq1`: This function creates the Lorenz curve, and computes the coefficient of variation (CV) and the Gini coefficient for group-wise data on population shares and mean income levels;

- `ineq2`: This function creates the Lorenz curve, and computes the coefficient of variation (CV) and the Gini coefficient for group-wise data on population and income shares;

- `ineq3`: This function creates the Lorenz curve, and computes the coefficient of variation (CV) and the Gini coefficient for individual-level data on income levels;

- `ineq4`: This function creates the Lorenz curve, and computes the coefficient of variation (CV) and the Gini coefficient for group-wise data on population and mean income levels.

## Installation

You can install the development version of ineq from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("dbasu-umass/ineq")
```

## Example

Let us work through an example to illustrate the functions in this package. The following data on the distribution of consumption expenditure in rural India in 2011 is taken from NSS KI 68/1.0 (Key Indicators of Household Consumer Expenditure in India, June 2013).

```{r example}
library(ineq)

# population fractile
w2011r <- c(0.05,0.05,rep(0.1,8),0.05,0.05)

## mean consumption expenditure by fractile
y2011r <- c(521.44,665.84,783.24,904.57,1017.80,
            1135.97,1266.08,1426.76,1645.36,2007.46,
            2556.33,4481.18)

# compute measure of inequality
myresults <- ineq1(y=y2011r, w=w2011r)

```

Now, let us look at the results.

```{r cars}
# Coefficient of variation
myresults$CV

# Gini coefficient
myresults$Gini

# cumulative population and income share
myresults$lzdata

```


The Lorenz curve.
```{r pressure, echo = FALSE}
plot(myresults$lzcurve)
```


